On Optimality of Finite Element Integration

### 2015-12-09 TOMS_Paper/assembly.tex
Minor fixes

Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that can characterize our expressions. Section~\ref{sec:sec:se-algo} presents the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

@@

\begin{Prop}
\label{prop:optimal-approach}
Consider a multilinear form and the corresponding finite element integration loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z = M \setminus P$. Assume that pre-evaluating different monomials does not result in identical tables. Then, given the memory Constraints~\ref{const:Le} and~\ref{const:TH}, the transformation algorithm produces a $\Lambda'$ which is quasi-optimal (as in Definition~\ref{def:mln-quasi-optimality}). 
\end{Prop}
\begin{proof}
Given a monomial $m$, $\theta(m)$ provides the cost/gain due to pre-evaluating $m$. $\theta(m)$ is \textit{not} an estimate\footnote{An estimate is used, instead, in~\citeN{Kirby:TC}. The estimate is actually for the whole form, rather than for its monomials.}, so $P$ (and $Z$) are constructed correctly.

Let $\Lambda_{ops}$ denote the operation count (additions, products) in $\Lambda$. Let $\Lambda'_I \subset \Lambda'$ be the set of all innermost loops produced by the transformation algorithm (all $L_k$ loops in Figure~\ref{code:loopnest-opt}). We need to show that there is no other synthesis $\Lambda''_{I}$ such that $\Lambda''_{I_{ops}} < \Lambda'_{I_{ops}}$. For this, we need to show that:
\begin{enumerate}
\item the coordination of pre-evaluation with sharing elimination is optimal. This boils down to show that
\begin{enumerate}
\item pre-evaluating any $Z_P: Z_P \subseteq Z$ would result $\Lambda''_{ops} > \Lambda'_{ops}$
\item not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops} > \Lambda'_{ops}$
\end{enumerate}
\item sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces loop nests that are at least quasi-optimal (again, in the sense of Definition~\ref{def:mln-quasi-optimality})
\end{enumerate}

We discuss each of these points individually

\begin{enumerate}
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two generic monomials $m_1, m_2 \in M$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. In a scenario in which $m_{T_{1}} \cap m_{T_{2}}$, we would have that at least one table would be assigned to the same temporary, with the transformation algorithm skipping a potential global optimum. By assumption, we neglect this possibility. The assumption is sustained by (i) the pathological nature of the scenario, (ii) the complexity inherent in handling it, and (iii) the fact that such a potential global optimum would probably be very close to the retrieved solution.
\item ...
\end{enumerate}
\item ...
\end{enumerate}
%
%A) The only corner situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$. If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim. Assumption 1) prevents this situation, so the statement is true. 
%
%B) Let $\theta_{wo}(m)$ and $\theta_{w}(m)$ represent, respectively, the operation count for a monomial $m$ without and with pre-evaluation and let us consider the non-trivial case in which two monomials $p \in P_Z$ and $z \in Z$ share at least one term. We have to contemplate the fact that despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle. That is, there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation. Potential examples are forms in which multiple monomials differ only for the number and/or the space of coefficient functions. The previously discussed assumption 2) allows us to neglect this possibility. 
\end{proof}

>>>>

Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that can characterize our expressions. Section~\ref{sec:se-algo} presents the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

@@

\begin{Prop}
\label{prop:optimal-approach}
Consider a multilinear form and the corresponding finite element integration loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z = M \setminus P$. Assume that pre-evaluating different monomials does not result in identical tables. Then, given the memory Constraints~\ref{const:Le} and~\ref{const:TH}, the transformation algorithm produces a $\Lambda'$ which is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}.
\end{Prop}
\begin{proof}
The function $\theta$ provides cost or gain of pre-evaluation. $\theta$ retrieves exact values, rather than estimates\footnote{An estimate is used, instead, in~\citeN{Kirby:TC}. The estimate is actually for the whole form, rather than for its monomials.}, so $P$ and $Z$ are constructed correctly.

Let $\Lambda_{ops}$ denote the operation count (number of additions and products) in $\Lambda$. Let $\Lambda'_I \subset \Lambda'$ be the set of all innermost loops produced by the transformation algorithm (all $L_k$ loops in Figure~\ref{code:loopnest-opt}). We need to show that there is no other synthesis $\Lambda''_{I}$ such that $\Lambda''_{I_{ops}} < \Lambda'_{I_{ops}}$, which holds if and only if
\begin{enumerate}
\item \textit{The coordination of pre-evaluation with sharing elimination is optimal}. This boils down to prove that
\begin{enumerate}
\item \textit{pre-evaluating any $Z_P: Z_P \subseteq Z$ would result $\Lambda''_{ops} > \Lambda'_{ops}$}
\item \textit{not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops} > \Lambda'_{ops}$}
\end{enumerate}
~
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two generic monomials $m_1, m_2 \in M$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. In a scenario in which $m_{T_{1}} \cap m_{T_{2}}$, we would have that at least one table would be assigned to the same temporary, with the transformation algorithm skipping a potential global optimum. By assumption, we neglect this possibility. The assumption is sustained by (i) the pathological nature of the scenario, (ii) the complexity inherent in handling it, and (iii) the fact that such a potential global optimum would probably be very close to the retrieved solution.
\item ...
\end{enumerate}

\item \textit{Sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces loop nests that are at least quasi-optimal.}\\
...
\end{enumerate}

%
%A) The only corner situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$. If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim. Assumption 1) prevents this situation, so the statement is true. 
%
%B) Let $\theta_{wo}(m)$ and $\theta_{w}(m)$ represent, respectively, the operation count for a monomial $m$ without and with pre-evaluation and let us consider the non-trivial case in which two monomials $p \in P_Z$ and $z \in Z$ share at least one term. We have to contemplate the fact that despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle. That is, there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation. Potential examples are forms in which multiple monomials differ only for the number and/or the space of coefficient functions. The previously discussed assumption 2) allows us to neglect this possibility. 
\end{proof}
